---
title: "How mizerShelf extends mizer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How mizerShelf extends mizer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(mizer)
library(mizerShelf)
params <- NWMed_params
```

# Overview

mizerShelf is an extension package for mizer that adds two extra ecosystem
components to a standard multispecies size-spectrum model:

- **detritus** — particulate organic matter that settles to the seafloor and
  provides food for small benthic organisms and juvenile fish,
- **carrion** — fresh animal tissue from natural deaths, fishing gear mortality,
  and discards that is consumed by scavengers before decomposing into detritus.

Both components are dynamical: their biomasses change over time in response to
species abundances and fishing effort. The package uses four of the five
extension mechanisms described in `vignette("extensions", package = "mizer")`:

| Extension mechanism | Used for |
|---|---|
| `setRateFunction()` | Replace standard mortality with shelf mortality that includes excess gear mortality |
| Resource dynamics override | Implement detritus dynamics in place of the default semi-chemostat resource |
| `setComponent()` | Add carrion as a scalar dynamical component that contributes to the encounter rate |
| S4 subclassing + S3 dispatch | Define `mizerShelf`/`mizerShelfSim` marker classes for shelf-specific plotting, scaling, and species manipulation |

All four mechanisms are wired together inside `newDetritusCarrionParams()`, which
is the single entry point for building a shelf model.

# The `mizerShelf` marker class

mizerShelf defines two S4 marker subclasses:

```r
setClass("mizerShelf",    contains = "MizerParams")
setClass("mizerShelfSim", contains = "MizerSim")
```

These classes carry no extra slots. Their purpose is to trigger S3 dispatch so
that shelf-specific behaviour is used automatically whenever a user calls a
mizer generic on a `mizerShelf` object. The classes are marker classes in the
sense described in `vignette("extensions", package = "mizer")`.

`newDetritusCarrionParams()` records the extension in `params@extensions` and
coerces the result to the `mizerShelf` marker class:

```r
params@extensions <- c(mizerShelf = as.character(packageVersion("mizerShelf")))
new("mizerShelf", params)
```

Recording the extension in `params@extensions` is what allows mizer to
automatically produce a `mizerShelfSim` whenever `project()` is called:
mizer's own `MizerSim()` constructor calls `coerceToExtensionClass()`, which
reads `params@extensions`, sees `mizerShelf`, and coerces the new sim object
to `mizerShelfSim`.

## Methods dispatched on `mizerShelf`

The following S3 methods are dispatched on `mizerShelf` or `mizerShelfSim`:

| Method | What it adds |
|---|---|
| `getBiomass.mizerShelf()` | Adds detritus and carrion biomasses to the species biomasses |
| `getBiomass.mizerShelfSim()` | Same, as a time series |
| `steady.mizerShelf()` | Calls `tune_carrion_detritus()` after convergence |
| `scaleModel.mizerShelf()` | Adjusts carrion encounter rates and external detritus input when rescaling abundances |
| `removeSpecies.mizerShelf()` | Updates the carrion encounter-rate matrix `rho` |
| `addSpecies.mizerShelf()` | Computes the initial `rho_carrion` for new species |

Each of these methods calls `NextMethod()` so that it first runs the standard
mizer behaviour and then makes only the shelf-specific adjustments.

### Example: `getBiomass`

The standard `getBiomass()` returns only species biomasses. The shelf version
appends the detritus and carrion biomasses:

```r
getBiomass.mizerShelf <- function(object, ...) {
    params <- object
    b <- NextMethod()                               # standard species biomasses
    d_biomass <- sum(params@initial_n_pp *
                     params@dw_full * params@w_full) # integrate resource spectrum
    b <- c(b, Detritus = d_biomass)
    scalar_other <- Filter(
        function(x) is.numeric(x) && length(x) == 1,
        params@initial_n_other)
    if (length(scalar_other) > 0) b <- c(b, unlist(scalar_other))
    b
}
```

Because `plotBiomass()` calls `getBiomass()` internally, this single override
makes biomass plots show detritus and carrion alongside fish species without any
further changes.

```{r getBiomass-example}
getBiomass(NWMed_params)[1:3]
```

```{r plotBiomass-example, fig.width=7, fig.height=4}
sim <- project(NWMed_params, t_max = 3, t_save = 0.5, progress_bar = FALSE)
plotBiomass(sim, species = c("Nut clam", "Detritus"))
```

# Detritus: overriding the resource dynamics

Mizer's standard resource spectrum follows a semi-chemostat towards a carrying
capacity. mizerShelf replaces this with `detritus_dynamics()`, which is
registered via the `resource_dynamics` argument to `newMultispeciesParams()`:

```r
params <- newMultispeciesParams(
    species_params = species_params,
    resource_dynamics = "detritus_dynamics",
    ...
)
```

This is the lightest-weight way to change the resource because it reuses all of
mizer's resource infrastructure — the size grid, the resource abundance array
`n_pp`, the `resource_mort` rate, and so on — while replacing only the update
step.

## The detritus ODE

The detritus spectrum is always held at a fixed power-law shape; only its
overall biomass $B$ is dynamical. The biomass satisfies

$$\frac{dB}{dt} = \text{production} - \text{consumption} \cdot B + \text{external}$$

where:

- **production** comes from unassimilated food (faeces), decomposing carrion,
  and an external input from the pelagic zone
  (`getDetritusProduction()`),
- **consumption** is the total rate at which species feed on detritus
  (`detritus_consumption()`), proportional to the current biomass,
- **external** is a tunable constant that represents sedimentation from
  pelagic waters, stored in `other_params(params)$detritus$external`.

`detritus_dynamics()` solves this ODE analytically within each time step to
avoid the instabilities of a simple Euler step:

$$B(t + \Delta t) = B(t)\,e^{-c\,\Delta t} + \frac{p}{c}\left(1 - e^{-c\,\Delta t}\right)$$

where $c$ is the mass-specific consumption rate and $p$ is the production rate.
The shape of the spectrum is then scaled to the new biomass:

```r
detritus_dynamics <- function(params, n, n_pp, n_other, rates, dt, ...) {
    current_biomass <- detritus_biomass(params, n_pp = n_pp)
    consumption     <- detritus_consumption(params, n_pp, rates) / current_biomass
    production      <- sum(getDetritusProduction(params, n, n_other, rates))
    if (consumption) {
        et           <- exp(-consumption * dt)
        next_biomass <- current_biomass * et + production / consumption * (1 - et)
    } else {
        next_biomass <- current_biomass + production * dt
    }
    n_pp * next_biomass / current_biomass
}
```

The function receives `n_pp` (the current resource spectrum), not the initial
spectrum, so the update is applied to whatever state the simulation is
currently in.

## Steady-state tuning

`tune_carrion_detritus()` adjusts the external detritus input so that at the
current abundances the detritus is at steady state (`dB/dt = 0`):

```r
params@other_params$detritus$external <- outflow - production
```

This function is called automatically by `steady.mizerShelf()` after mizer's
own steady-state iteration converges.

## User-facing controls

Two getter/setter pairs make it easy to calibrate the detritus:

- `detritus_biomass(params)` — total detritus biomass in grams.
- `detritus_lifetime(params)` — the expected time a unit of detritus survives
  before being consumed, equal to $B / (\text{consumption rate})$.
- `detritus_lifetime(params) <- value` — rescales detritus abundance while
  keeping total consumption unchanged (by adjusting `interaction_resource`).

```{r detritus-lifetime}
detritus_lifetime(params)
```

# Carrion: adding a new dynamical component

Carrion is not naturally represented by any existing mizer structure: it is a
scalar total biomass (not size-structured), produced by mortality and consumed
by scavengers with a species- and size-dependent encounter rate. It is
registered as an *other* component via `setComponent()`:

```r
params <- setComponent(
    params, "carrion",
    initial_value  = 1,
    dynamics_fun   = "carrion_dynamics",
    encounter_fun  = "encounter_contribution",
    component_params = list(rho = rho))
```

The three arguments name the functions that mizer will call at each time step:

- `dynamics_fun = "carrion_dynamics"` — updates the carrion biomass.
- `encounter_fun = "encounter_contribution"` — adds the carrion's contribution
  to each species' encounter rate.
- `component_params` stores the `rho` matrix (species × size) and a
  decomposition rate in `other_params(params)$carrion`.

## Encounter contribution

The contribution of carrion to the encounter rate of species $i$ at body size
$w$ is

$$E_i^{\text{carrion}}(w) = \rho_i(w)\,B_{\text{carrion}}$$

where $\rho_i(w)$ is a species- and size-dependent encounter rate coefficient
stored in the component parameters. The function `encounter_contribution()` is
a generic helper that works for any scalar other-component:

```r
encounter_contribution <- function(params, n_other, component, ...) {
    params@other_params[[component]]$rho * n_other[[component]]
}
```

The `rho` matrix is initialised in `newDetritusCarrionParams()` so that each
species has enough food from all sources (fish prey, detritus, and carrion) to
reach a feeding level of `f0` at maximum body size.

## Carrion dynamics

The carrion biomass $B$ satisfies the same ODE structure as the detritus:

$$\frac{dB}{dt} = \text{production} - \text{consumption} \cdot B$$

- **production** (`getCarrionProduction()`) comes from three sources:
  - a fraction `ext_prop` of natural (non-predation) mortality,
  - excess gear mortality (`gearMort()`), i.e. animals killed by the gear
    but not landed,
  - discards — landed animals that are thrown back.
- **consumption** (`carrion_consumption_ms()`) includes active consumption by
  fish scavenging carrion plus bacterial decomposition.

As with detritus, the ODE is solved analytically within each time step:

```r
carrion_dynamics <- function(params, n, n_other, rates, dt, ...) {
    consumption <- carrion_consumption_ms(params, n, rates)
    production  <- sum(getCarrionProduction(params, n, rates))
    if (consumption) {
        et <- exp(-consumption * dt)
        return(n_other$carrion * et + production / consumption * (1 - et))
    }
    return(n_other$carrion + production * dt)
}
```

## Production and consumption breakdown

The package provides functions to inspect where carrion comes from and where it
goes:

```{r carrion-production, fig.width=5, fig.height=5}
plotCarrionProduction(params)
```

```{r carrion-consumption, fig.width=5, fig.height=5}
plotCarrionConsumption(params)
```

## User-facing controls

- `carrion_lifetime(params)` — expected lifespan of carrion in years.
- `carrion_lifetime(params) <- value` — rescales the carrion biomass and its
  encounter-rate coefficients while keeping total consumption unchanged.
- `carrion_human_origin(params)` — fraction of carrion production that comes
  from fishing (gear mortality + discards).
- `carrion_human_origin(params) <- value` — adjusts `ext_prop` (the fraction
  of natural mortality that produces carrion) to hit the target.

```{r carrion-lifetime}
carrion_lifetime(params)
carrion_human_origin(params)
```

# The custom mortality rate

Standard mizer fishing mortality accounts only for fish that are extracted from
the water. Continental shelf fisheries also kill animals that remain on the
seabed after contact with bottom-trawl gear. mizerShelf captures this
distinction with a custom mortality function registered via `setRateFunction()`:

```r
params <- setRateFunction(params, "Mort", "seMort")
```

`seMort()` calls the standard `mizerMort()` and adds the excess gear mortality
on top:

```r
seMort <- function(params, n, n_pp, n_other, t, f_mort, pred_mort, ...) {
    mizerMort(params, n, n_pp, n_other, t, f_mort, pred_mort, ...) +
        gearMort(params, f_mort = f_mort)
}
```

`gearMort()` returns `max(gear_mort - f_mort, 0)`, where `gear_mort` is a
species parameter giving the total rate at which animals die due to gear
contact. When fishing mortality is below this cap, the remainder enters total
mortality and also enters the carrion production (via `getCarrionProduction()`).
When fishing mortality exceeds `gear_mort`, excess gear mortality is zero.

The custom rate function receives and returns exactly the same array shape as
`mizerMort()` (species × size), so it is a drop-in replacement that adds
one ecological process without changing anything else.

# How the pieces fit together in `newDetritusCarrionParams()`

All four extension mechanisms are wired up inside a single constructor:

```r
newDetritusCarrionParams <- function(species_params,
                                     w_min_detritus = NA,
                                     w_max_detritus = 1,
                                     n = 0.7, ...) {
    # 1. Detritus: override resource dynamics
    params <- newMultispeciesParams(
        species_params   = species_params,
        min_w_pp         = w_min_detritus,
        w_pp_cutoff      = w_max_detritus,
        n = n, p = n,
        resource_dynamics = "detritus_dynamics",
        ...)

    # Initialise rho so each species can reach feeding level f0 at max size
    f0  <- set_species_param_default(params@species_params, "f0", 0.6)$f0
    ic  <- set_species_param_default(
               params@species_params, "interaction_carrion", 1)$interaction_carrion
    E   <- getEncounter(params)[, length(params@w)] /
               (params@w[length(params@w)] ^ n)
    rho <- pmax(0, f0 * params@species_params$h / (1 - f0) - E) * ic
    params@species_params$rho_carrion <- rho
    rho <- outer(params@species_params$rho_carrion, params@w ^ n)

    # 2. Mortality: custom rate function adds excess gear mortality
    params <- setRateFunction(params, "Mort", "seMort")

    # 3. Carrion: scalar dynamical component
    params <- setComponent(
        params, "carrion", initial_value = 1,
        dynamics_fun     = "carrion_dynamics",
        encounter_fun    = "encounter_contribution",
        component_params = list(rho = rho))

    # 4. Marker class: enables S3 dispatch for shelf-specific methods
    params <- setColours(params,
                         c(Detritus = "forestgreen", carrion = "peru"))
    new("mizerShelf", params)
}
```

The order matters. The resource dynamics are set inside `newMultispeciesParams()`
before mizer computes initial rates, so the detritus shape is consistent from
the start. The carrion component is registered after the species parameters are
set, so that `rho` can be computed from the initial encounter rates. The marker
class is applied last, after all the standard mizer slots have been populated.

# Summary

mizerShelf illustrates all four of the core mizer extension mechanisms working
together:

1. **Resource dynamics override** — detritus reuses mizer's resource spectrum
   infrastructure with a custom ODE that tracks total biomass and rescales the
   spectrum shape.

2. **`setRateFunction()`** — the custom `seMort()` adds excess gear mortality to
   the standard mortality calculation with a single-line wrapper around
   `mizerMort()`.

3. **`setComponent()`** — carrion is a scalar dynamical variable whose biomass
   evolves each time step and whose encounter-rate contribution is automatically
   included in `getEncounter()` and `getDiet()`.

4. **S4 marker classes + S3 dispatch** — `mizerShelf` and `mizerShelfSim`
   ensure that shelf-specific methods (`getBiomass`, `steady`, `scaleModel`,
   `addSpecies`, `removeSpecies`) are dispatched automatically, while every
   override delegates to `NextMethod()` so that the full mizer pipeline remains
   intact.

Together, these mechanisms allow mizerShelf to add two ecologically important
processes — benthic detritus and carrion scavenging — to any mizer model
without modifying the mizer source code.
