---
title: "Model Description"
output:
  html_document:
    toc: yes
    toc_depth: 3
    code_folding: hide
  pdf_document:
    toc: yes
    toc_depth: '3'
vignette: |
  %\VignetteIndexEntry{Model Description} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(mizerNWMed)
library(ggplot2)
library(dplyr)
library(knitr)
```

In this document we present the details of the model. The model consists of three
components: 1) a collection of species modelled by [size spectrum dynamics],
2) [carrion] and 3) [detritus]. We discuss each of these in turn below.

In addition to describing the general model, we also give the values of the
parameters for the model tuned to the shelf ecosystem off the coast of Blanes
in the Northwestern Mediterranean. These are by default hidden in collapsible
sections and can be
unhidden by clicking on the "Parameter values" headings as below:
<details>
<summary>Parameter values</summary>
And if you click again this section will be hidden again.
</details>


# Size spectrum dynamics

## Species size spectra
The model assumes that, to a first approximation, an individual can be
characterized by its weight $w$ and its species number $i$ only. The aim of the
model is to calculate the size spectrum $N_i(w)$, which is the *density* of
individuals of species $i$ such that $\int_w^{w+dw}N_i(w)dw$ is the *number* of
individuals of species $i$ in the size interval $[w,w+dw]$. In other words: the
number of individuals in a size range is the area under the number density
$N_i(w)$.

The time evolution of the number density $N_i(w)$ is described by the 
McKendrick-von Foerster equation, which is a transport equation describing
the transport of biomass from small to large individuals,
with an additional loss term due to fish mortality:

\begin{equation}
  \label{eq:MvF}
  \frac{\partial N_i(w)}{\partial t} + \frac{\partial g_i(w) N_i(w)}{\partial w} 
  = -\mu_i(w) N_i(w).
\end{equation}

The individual growth rate $g_i(w)$ is described below in the [Growth](#growth)
section and the mortality rate $\mu_i(w)$ is described in the [Mortality](#mortality)
section. These rates
depend on the density of other fish of other sizes, as well as the carrion
and detritus biomasses, making the size-spectrum
dynamics non-linear and non-local in very interesting ways. The resulting
effects are too complicated to disentangle by pure thought. This is where
simulations with the mizer package come in.

There is no need to understand the mathematical notation used in
the McKendrick-von Foerster equation to understand its origin: it just says that the rate at which the
number of fish in a size bracket changes is the rate at which fish grow into the
size bracket from a smaller size minus the rate at which fish grow out of it to
a larger size minus the rate at which the fish in the size bracket die. 

Of course there also needs to be reproduction into the smallest size class,
otherwise there would be no small fish any more after a while. So for the
smallest size class, instead of a rate of growth into the size class there is
a rate of reproduction of new individuals into that size class. This
reproduction will be described below in the [Reproduction](#reproduction)
section.



## Growth

Consumers can grow only by consuming prey (including possibly carrion and
detritus). In the next few subsections we will
build towards determining the growth rate resulting from predation.
We will discuss how we model the [predator-prey encounter rate], the resulting rate
of [consumption], the rate of [metabolic losses], and the partitioning of the
remaining energy into [reproduction](#sec:repro) and [somatic growth](#somatic-growth).

### Predator-prey encounter rate {#sec:pref}

The rate $E_{i}(w)$ at which a predator of species $i$ and weight $w$ 
encounters food  (mass per time) is
determined by summing over all prey species and
integrating over all prey sizes $w_p$, weighted by the selectivity factors,
and (where relevant) adding the encounter rates $E_{C.i}$ of carrion and $E_{D.i}$ of detritus:
\begin{equation}
  \label{eq:1}
  E_{i}(w) = \gamma_i(w) \int \sum_{j} \theta_{ij} N_j(w_p)
  \phi_i(w,w_p) w_p \, dw_p + E_{i.C}(w) + E_{i.D}(w).
\end{equation}
The encounter rates for
[carrion](#carrion-consumption) and [detritus](#detritus-consumption) will be described later.

The overall prefactor $\gamma_i(w)$ sets the predation power of the predator. It
could be interpreted as a search volume or as an attack rate. By 
default it is assumed to scale allometrically as
$\gamma_i(w) = \gamma_i\, w^{3/4}.$ In order for $E_i(w)$ to have units of
gramms per year, the prefactor $\gamma_i$ has to have a unit of 
$\rm{gramms}^{-3/4}$ per year.

The $\theta_{ij}$ matrix sets the interaction strength between predator
species $i$ prey species $j$.

The size selectivity is encoded in the predation kernel $\phi_i(w,w_p)$.
For most predator species we use the lognormal predation kernel
given as
$$\phi_i(w, w_p) = 
\exp \left[ \frac{-(\ln(w / w_p / \beta_i))^2}{2\sigma_i^2} \right]
$$
if $w/w_p$ is larger than 1 and zero otherwise.
Here $\beta_i$ is the preferred predator-prey mass ratio and $\sigma_i$
determines the width of the kernel.

For Gurnards, Hake, Angler fish, Poor cod and Blue whiting we use a power-law
kernel with sigmoidal cutoffs given by 
$$\phi_i(w, w_p) = 
\frac{(w/w_p)^s}{\left(1+e^{l_l}\frac{w_p}{w}\right)^{u_l}
\left(1+e^{-l_r}\frac{w}{w_p}\right)^{u_r}}.
$$

Here the parameters $l_l$ and $u_l$ determine the sigmoidal cutoff at low
predator/prey mass ratio and $l_r$ and $u_r$ similarly determine the cutoff
at large predator/prey mass ratio.

The parameters for these predation kernels were estimated from stomach data
or from the physical characteristics of the species. This will be discussed
elsewhere.

<details>
<summary>Parameter values</summary>

For the species that use a lognormal predation kernel, the parameters are
```{r results='asis'}
params@species_params |>
    filter(pred_kernel_type == "lognormal") |>
    select(beta, sigma, gamma) |>
    kable()
```

For the species that use a truncated power law predation kernel, the parameters are
```{r results='asis'}
params@species_params |>
    filter(pred_kernel_type == "power_law") |>
    select(starts_with("kernel"), gamma) |>
    rename(s = kernel_exp,
           l_l = kernel_l_l,
           u_l = kernel_u_l,
           l_r = kernel_l_r,
           u_r = kernel_u_r) |>
    kable()
```

The predator/prey interaction matrix has most entries equal to either 0
(if the species can not interact) or 1. Only the interaction between Hake
as predator and Gurnards as prey is set to 0.5.

```{r}
ggplot(melt(params@interaction),
       aes(x = prey, y = predator, fill = value)) + 
  geom_tile(colour = "black", size = 0.25) +
    scale_fill_gradient(low = "white", high = "blue", limit = c(0, 1),
                        guide = "legend") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```


</details>

### Consumption

The encountered food is consumed subject to a standard Holling functional 
response type II to represent satiation. This determines the 
*feeding level* $f_i(w)$, which is a dimensionless number between 0 
(no food) and 1 (fully satiated) so that $1-f_i(w)$ is the proportion of the
encountered food that is consumed. The feeding level is given by

\begin{equation}
  \label{eq:f}
  f_i(w) = \frac{E_i(w)}{E_i(w) + h_i(w)},
\end{equation}

where $h_i(w)$ is the maximum consumption rate of a predator of species $i$
and weight $w$. By default we assume an allometric form
$h_i(w) = h_i\, w^n$
with $n=0.7$. The unit of the coefficients $h_i$ are $\rm{gramms}^{1-n}$ per year.

The rate at which food is consumed by a predator of species $i$ and weight $w$
is then 
\begin{equation}
(1-f_i(w))E_{i}(w)=f_i(w)\, h_i(w).
\end{equation}
Only a proportion $\alpha_i$ of this consumed biomass is retained, while a
proportion $1-\alpha_i$ is expelled in the form of feces, which contribute to
the detritus.

<details>
<summary>Parameter values</summary>
The values for the coefficients $h_i$ in the maximum consumption rates
were chosen so that the feeding level that fish experience has a reasonable
value with fish being neiter too starved nor totally satiated.
```{r}
params@species_params |>
    select(h, alpha, n) |>
    kable()
```

```{r}
plotFeedingLevel(params)
```

</details>

### Metabolic losses

Some of the consumed food is used to fuel the needs for metabolism and 
activity and movement, at a rate ${\tt metab}_i(w)$. By default 
this is made up out of standard metabolism, scaling with exponent $p$, and
loss due to activity and movement, scaling with exponent $1$:
\[{\tt metab}_i(w) = k_{s.i}\,w^p + k_i\,w.\]
The units of the coefficients $k_{s.i}$ are $\rm{gramms}^{1-p}$ per year and
the units of the $k_i$ is gramms per year.

The remaining energy, if any, 
is then available for growth and reproduction, at the rate
\begin{equation}
  \label{eq:Er}
  E_{r.i}(w) = \max(0, \alpha_i f_i(w)\, h_i(w) - {\tt metab}_i(w))
\end{equation}

<details>
<summary>Parameter values</summary>
```{r}
params@species_params |>
    select(ks, p, k) |>
    kable()
```
</details>

### Investment into reproduction  {#sec:repro}
A proportion $\psi_i(w)$ of the energy available for growth and reproduction is
used for reproduction. This proportion changes from zero below the weight
$w_{m.i}$ of maturation to one at the asymptotic weight $w_{\infty.i}$, where
all available energy is used for reproduction. The expression is
$$ \psi_i(w) = \begin{cases}
\left[1+\left(\frac{w}{w_{mat}}\right)^{-U}\right]^{-1}
\left(\frac{w}{w_{inf}}\right)^{m-n}&w<w_{inf}\\
1&w\geq w_{inf}\end{cases}
$$

with $m-n = 0.3$ and $U=10$ (which sets the steepness of the sigmoidal
switch-on of reproduction at around the maturity weight $w_{mat}$).
See
`mizer::setReproduction()` for more detail. 

<details>
<summary>Parameter values</summary>
```{r}
params@species_params |>
    select(w_mat, w_inf) |>
    kable()
```
</details>

### Somatic growth
What is left over after metabolism and reproduction is taken into account
is invested in somatic growth. Thus the growth rate of an individual of
species $i$ and weight $w$ is
\begin{equation}
  \label{eq:growth}
  g_i(w) = E_{r.i}(w)\left(1-\psi_i(w)\right).
\end{equation}
When food supply does not cover the requirements of metabolism and activity, 
growth and reproduction stops, i.e. there is no negative growth.

<details>
<summary>Parameter values</summary>
The values for the model parameters were chosen so that the resulting 
growth curves would be close to von Bertalanffy growth curves with the 
following parameters that were taken from the literature:

```{r}
params@species_params |>
    select(k_vb, t0, a, b) |>
    kable()
```
Here the parameters $a$ and $b$ are parameters for the allometric weight-length
relationship $w = a l^b$ where $w$ is measured in gramms and $l$ is measured
in centimeters.

```{r, out.width="100%"}
plotGrowthCurves(params, species_panel = TRUE)
```

</details>

## Mortality
The mortality rate $\mu_i(w)$ of an individual has four sources: 
predation mortality $\mu_{p.i}(w)$, background mortality $\mu_{ext.i}(w)$, 
fishing mortality $\mu_{f.i}(w)$ and excess gear mortality $\mu_{g.i}$, 
\[\mu_i(w)=\mu_{p.i}(w)+\mu_{ext,i}(w)+\mu_{f.i}(w)+\mu_{g.i}(w).\]
We will now explain each of the terms.

### Predation mortality
All consumption by fish translates into 
corresponding predation mortalities on the ingested prey individuals. 
Recalling that $1-f_j(w)$ is the proportion of the food encountered by a 
predator of species $j$ and weight $w$ that is actually consumed, the
rate at which all predators of species $j$ consume prey of size $w_p$ is
\begin{equation}
  \label{eq:pred_rated}
  {\tt pred\_rate}_j(w_p) = \int \phi_j(w,w_p) (1-f_j(w))
  \gamma_j(w) N_j(w) \, dw.
\end{equation}
This predation rate is calculated by the function `getPredRate()`.

The mortality rate due to predation is then obtained as
\begin{equation}
  \label{eq:mup}
  \mu_{p.i}(w_p) = \sum_j {\tt pred\_rate}_j(w_p)\, \theta_{ji}.
\end{equation}


### External mortality
External mortality $\mu_{ext.i}(w)$ is independent of the abundances. By default mizer assumes that the external
mortality for each species is a constant $z0_i$ independent of size. The value
of $z0_i$ is either specified as a species parameter or it is assumed to
depend allometrically on the asymptotic size:
\[z0_i = z0_{pre} w_{\infty.i}^{1-n}.\]

<details>
<summary>Parameter values</summary>
We use size-independent external mortalities.
```{r}
params@species_params |>
    select(z0) |>
    kable()
```
</details>

### Fishing mortality
Fishing mortality $\mu_{f.i}(w)$ is calculated with the function `mizer::getFMort()`.
The fishing parameters for the model are set up with `mizer::setFishing()`, where
you can find the details of how to set up gears with different selectivities
and the capabilities of different species.

<details>
<summary>Parameter values</summary>
TODO: write some explanations
```{r}
params@gear_params |>
    filter(sel_func == "knife_edge") |>
    select(species, knife_edge_size, catchability) |>
    kable(row.names = FALSE)
```

```{r}
params@gear_params |>
    filter(sel_func == "sigmoid_length") |>
    select(species, l50, l25, catchability) |>
    kable(row.names = FALSE)
```

</details>

### Excess gear mortality
The fishing mortality only includes individuals that are hauled into the
fishing vessel. Fishing gear also causes mortality among individuals that
encounter the gear but are not caught by it. This mortality is assumed not
to be size-specific.
There is a species parameter called `gear_mort` that gives the 
mortality rate of an individual imposed by the fishing gear. The part of
this gear mortality that is not fishing mortality (i.e., the part where the
individuals are not taken up to the fishing vessel but left dead in the sea)
we denote as the excess gear mortality.
$$\mu_{g.i} = \max{\left(\rm{gear\_mort}_i - \mu_{f.i}(w), 0\right)}
$$

This excess gear mortality contributes to the carrion production.

<details>
<summary>Parameter values</summary>
```{r}
params@species_params |>
    select(gear_mort) |>
    kable()
```
</details>

## Reproduction

### Energy invested into reproduction

The total rate of investment into reproduction (grams/year) is
found by integrating the contribution from all individuals of species $i$,
each of which invests a proportion $\psi_i(w)$ of their consumption. This
total rate of energy investment can then be converted to a rate of
production of offspring $R_{p.i}$ (numbers per year):
\begin{equation}
  \label{eq:Rp}
  R_{p.i} = \frac{\epsilon}{2 w_{min}} \int N_i(w)  E_{r.i}(w) \psi_i(w) \, dw,
\end{equation}
Here the total rate of investment is multiplied by an efficiency factor $\epsilon$ 
and then dividing by the offspring weight $w_{min}$ to convert the energy into number of offspring.
The result is multiplied by a factor $1/2$ to take into account that only 
females reproduce.

Note that for species that have a pelagic phase the size $w_min$ is the size at
which the offspring join the benthic ecosystem.



### Density-dependence in reproduction

Three important density-dependent mechanisms widely assumed in fisheries models are 
automatically captured in the mizer model that lead to an emergent stock-recruitment relationship:

1. High density of spawners leads to a reduced
food income of the spawners and consequently reduced per-capita reproduction.
2. High density of larvae leads to slower growth of larvae due to food 
competition, exposing the larvae to high mortality for a longer time, thereby
decreasing the survivorship to recruitment size.
3. High density of fish leads to more predation on eggs and fish larvae by
other fish species or by cannibalism.

However there are other sources of density dependence that are not explicitly
modelled mechanistically in mizer. An example would be a limited carrying capacity of suitable
spawning grounds and other spatial effects.

This requires additional phenomenological density dependent contributions to the stock-recruitment. In mizer this type of density dependence is modelled through constraints on egg production and survival. 
The default functional form of this density dependence is represented by
a reproduction rate $R_i$ (numbers per time) that approaches a maximum as the 
energy invested in reproduction increases, modelled mathematically it is analogous to 
a *Beverton-Holt* type function:

\begin{equation}
  \label{eq:R}
  R_i = R_{\max.i} \frac{R_{p.i}}{R_{p.i} + R_{\max.i}},
\end{equation}
where $R_{\max.i}$ is the maximum reproduction rate of species $i$.

<details>
<summary>Parameter values</summary>
```{r}
params@species_params |>
    select(w_min, erepro, R_max) |>
    kable()
```
</details>

# Carrion

Carrion (consisting of the dead individuals that have not yet decomposed) is an important component of the ecosystem,
providing food for scavenger species. Feeding on carrion by scavengers
is not size-based. Scavengers can feed on carrion of any size. Therefore
we do not need to describe the carrion by a size spectrum but only need
to describe its total biomass $B_C$.

The rate of change in the total carrion biomass is simply the difference
between the rate at which carrion biomass is produced and the rate at which
it is consumed, so
$$\frac{dB_C}{dt}=p_C - c_C\,B_C.$$

We will discuss the production rate $p_C$ and the consumption rate
$c_CB_C$ below.

## Carrion production
The rate $p_C$ at which carrion biomass is produced by the ecosystem has
contributions from three sources, 
$$p_C = p_{C.ext} + p_{C.g} + p_{C.d},$$
each of which we will now discuss.

### External mortality
$p_{C.ext}$ comes from animals that have died by natural causes other than predation ("external"):
  A mizer model allows for external mortality to describe all deaths by
  natural causes that are not due to predation from the modelled species.
  So this external mortality would include deaths that lead to carrion,
  but also deaths due to predation by species that
  are not explicitly modelled, for example mammals or sea birds. Thus only
  a proportion of the external mortality produces carrion.
  This is given by a carrion parameter `ext_prop`. So
  $$p_{C.ext} = \rm{ext\_prop}\,\sum_i\int\mu_{ext.i}(w)N_i(w)w\,dw.$$

<details>
<summary>Parameter values</summary>
The value of `ext_prop` is `r params@other_params$carrion$ext_prop`.
</details>

### Excess gear mortality
$p_{C.g}$ comes from animals killed by the fishing gear that are not taken 
up to the fishing vessel but left dead in the sea. Thus
$$p_{C.g} = \sum_i\int\mu_{g.i}N_i(w)w\,dw,$$
  where the excess gear mortality rate $\mu_{g.i}$ was discussed 
[above](#excess-gear-mortality).


### Discards
$p_{C.d}$ comes from discarding of fished animals ("discards"):
  There is a species parameter $d_i$, called `discard`, that gives the proportion of
  the catch biomass that is discarded. This biomass is added to the carrion
  biomass. Thus
  $$p_{C.d} = \sum_i\,d_i\int\mu_{f.i}(w)N_i(w)w\,dw.$$

<details>
<summary>Parameter values</summary>
```{r}
params@species_params |>
    select(discard) |>
    kable()
```
</details>

## Carrion consumption
Carrion is consumed by scavengers, but also decomposed by bacteria and other
processes. The rate at which carrion biomass is consumed is assumed to be proportional to the
available carrion biomass. The proportionality factor $c_C$, which we refer to
as the "mass-specific consumption rate", has one component that depends on the
abundance of consumers and a constant component $d_C$ representing decomposition.

For each consumer species $i$, a parameter $\rho_i$ determines the rate at which individuals
of that species encounter carrion biomass. The rate is assumed to
scale with the size of the predator raised to an allometric exponent $n$
which is taken to be the same as the scaling exponent of the maximum intake rate
for consumers, 
$$E_{i.C}(w)=\rho_i\, w^n\,B_C.$$
Finally, satiation of the consumers is taken into account
via their feeding level $f_i(w)$ that was described in the section on 
[consumption](#consumption). This gives the mass-specific carrion
consumption rate
$$c_C = \sum_i\int\rho_i\, w^n N_i(w) (1-f_i(w))\,dw + d_C$$
where $d_C$ is the mass-specific rate of decomposition.

<details>
<summary>Parameter values</summary>
The value of the mass-specific rate of decomposition is $d_C=$
`r params@other_params$carrion$decompose` per year. 

The parameters
$\rho_i$ have units of $g^{-n}$ per year. They are non-zero only for
species that do at least some scavenging:
```{r}
params@species_params |>
    select(rho_carrion) |>
    rename(rho = rho_carrion) |>
    filter(rho > 0) |>
    kable()
```
</details>

# Detritus

Detritus is at the base of the benthic foodweb, providing food for
detritivores. 

We describe the detritus as a size-spectrum $N_D(w)$ ,
which is the *density* of detritus particles of size $w$, 
so that $\int_w^{w+dw}N_D(w)dw$ is the *number* of
detritus particles in the size interval $[w,w+dw]$.
However, we do not know details about the size-specific dynamics
of detritus and simply assume that its abundance is described by
a power law up to a maximum size,
$$N_D(w) \propto \begin{cases} w^{-\lambda} & w\leq w_{cutoff}\\
0 & w > w_{cutoff}\end{cases}.
$$

The exponent $\lambda$ is kept fixed and only the coefficient of the
power law changes with time to reflect the change in the total detritus
biomass 
$$B_D = \int_{w_{min}}^{w_{cutoff}} N_D(w)\, w \,dw.$$

The rate of change in the total detritus biomass is simply the difference
between the rate at which detritus biomass is produced and the rate at which
it is consumed, so
$$\frac{dB_D}{dt}=p_D - c_D\,B_D.$$

We will discuss the production rate $p_D$ and the consumption rate
$c_DB_D$ below.

## Detritus production
The rate $p_C$ at which carrion biomass is produced by the ecosystem has
contributions from three sources, 
$$p_C = p_{C.f} + p_{C.c} + p_{C.ext},$$
each of which we will now discuss.

### Feces
$p_{C.f}$ comes from the biomass that is consumed but not assimilated by
the predators, i.e., it comes from the feces expelled by the predators.
Let $\alpha_i$ be the proportion of the consumed biomass that is assimilated
by species $i$
and let $f_i(w)$ be the feeding level and $E_i(w)$ the food encounter rate
discussed in the section on [consumption](#consumption). Then
$$p_{C.f} = \sum_i(1-\alpha_i)\int (1-f_i(w))E_i(w)\,dw.$$

### Decomposing carrion
$p_{D.c}$ comes from decomposing carrion. As we discussed in the section on
[carrion consumption], carrion biomass is decomposed to detritus at the rate
$d_CB_C$ where $d_C$ is a given fixed mass-specific decomposition rate and
$B_C$ is the total carrion biomass. So
$$p_{D.c}=d_CB_C.$$

### External
$p_{D.ext}$ is the rate at which detritus enters the system from external 
sources. This will mostly be detritus sinking in from the pelagic zone. 
This rate is a model parameter independent of any other model component.

<details>
<summary>Parameter values</summary>
The value of the external detritus production rate is $p_{d.ext}=$
`r params@other_params$detritus$external` grams per year. 

</details>

## Detritus consumption
The rate at which detritus biomass is consumed is assumed to be proportional to the
available detritus biomass. The proportionality factor $c_D$, which we refer to
as the "mass-specific consumption rate", depends on the
abundance of consumers.

The consumption of detritus is modelled similarly to the consumption of fish. First we introduce the rate at which
all predators of species $j$ consume detritus particles of size $w$
$$\mu_{D}(w_p) = \sum_j  \theta_{jD}\int \phi_j(w,w_p) (1-f_j(w))
  \gamma_j(w) N_j(w) \, dw.
$$
This is analogous to the [predation mortality] discussed earlier, but
with $\theta_{jD}$ determining the strength at which species $j$ feeds on
detritus. To get
the total rate of detritus consumption we multiply by the weight of the
detritus particle and integrate over all detritus particles:
$$c_DB_D=\int\mu_D(w_p)\,w_p\, N_D(w_p)\,dw_p.
$$
Because we keep the size-distribution of the detritus fixed, this
consumption rate is proportional to the total detritus biomass $B_D$, as we
have already indicated by our notation.

<details>
<summary>Parameter values</summary>
We use the same value of 
$\theta_{jD}$=`r params@species_params$interaction_resource[[1]]`
for all predator species.
</details>
